---
title: "Lecture 30: Lab session"
author: "Dr. Xiang Ji @ Tulane University"
date: "April 9, 2021"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-6040/7260 Linear Models
csl: ../apa.csl
---

## Reference

[ANCOVA example blog](https://www.r-bloggers.com/2019/04/ancova-example-april-18-2019/)


## ANCOVA example

Let's consider the `diamonds` dataset from `ggplot2` package.

Let’s say for the sake of argument you’d like to know more about how these factors of `cut` and `color` impact the `price` you’ll pay.
Let’s go ahead and get things set up in R so we can proceed, load the right libraries etc.. 

```{r}
library(tidyverse)
str(diamonds)
```

We follow the blog to process the data into a balanced design.

```{r}
set.seed(1234)
diamonds2 <- filter(diamonds, 
                    cut %in% c("Fair", "Good") & 
                    color %in% c("E", "F", "G") & 
                    carat < 1.75) 
diamonds2 <- droplevels(diamonds2)
one <- diamonds2 %>% filter(cut == "Fair" & color == "E") %>% sample_n(218)
two <- diamonds2 %>% filter(cut == "Fair" & color == "F") %>% sample_n(218)
three <- diamonds2 %>% filter(cut == "Fair" & color == "G") %>% sample_n(218)
four <- diamonds2 %>% filter(cut == "Good" & color == "E") %>% sample_n(218)
five <- diamonds2 %>% filter(cut == "Good" & color == "F") %>% sample_n(218)
six <- diamonds2 %>% filter(cut == "Good" & color == "G") %>% sample_n(218)
diamonds2 <- bind_rows(one, two, three, four, five, six)
str(diamonds2)
```

### Two-way ANOVA model

Now let's do a two-way ANOVA with `price` as the response and a cross design over `cut` and `color` (in other words, consider both the main effects of `cut` and `color` and their interaction effect).

Print out your ANOVA table


#### Is the interaction effect significant at $\alpha=0.05$ level?


#### Is the main-effect `cut` significant at $\alpha=0.05$ level?

#### Is the main-effect `color` significant at $\alpha=0.05$ level?


### Just looking at `carat`

Based on common sense, we want to control over the weight of the diamond by including `carat` as a covariate into our model.

Before fitting the model, let's simply look at the relationship between `price` and `carat` by making a plot.

```{r}
ggplot(diamonds2, aes(x = carat, y= price)) + 
  geom_point(alpha = 1/10) + 
  geom_smooth(method = "loess", color = "red") + 
  geom_smooth(method = "lm", color = "blue")
```

#### What is the blue line in the above plot?  What is the red line in the plot?


### Adding `carat` into the model

Use `price` as the response variable, `cut` and `color` as two main-effects with their interaction, and `carat` as a covariate.
Print out the ANOVA table (use the `Anova` function from the `car` package with argument `type = 2` to get type II sum of squares).

```{r, eval = FALSE}
two.way.ancova <- aov()
car::Anova(two.way.ancova, type = 2)
```

#### Are any of the effects significant now? Why?

